---
title: "Allele Specific Expression Analysis"
author: "Lachlan Baer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
    fig_align: "center"
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  fig.align = "center"
)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

# Setup 

```{r packages}
library(tidyverse)
library(magrittr)
library(parallel)
library(here)
library(AnnotationHub)
library(purrr)
library(scales)
library(kableExtra)
library(statmod)
```

```{r options}
if (interactive()) setwd(here::here())
theme_set(theme_bw())
cores <- detectCores() - 1
```

```{r}
source("~/bioinformatics/bioToolkit/lbFuncs.R")
```

## EnsDb

```{r}
ah <- AnnotationHub() %>%
  subset(species == "Danio rerio") %>%
  subset(rdataclass == "EnsDb")
ensDb <- ah[["AH83189"]] ## Ens101
genes <- genes(ensDb)
mcols(genes) <- mcols(genes)[
  c("gene_id", "gene_name", "gene_biotype", "entrezid")
]
exons <- exonsBy(ensDb, by = "gene")
```

# Load data

## Metadata

```{r}
metadata <- read_csv(here("files/samples.csv")) %>%
  dplyr::select(-sample) %>%
  dplyr::rename(sample = basename, genotype = Genotype) %>%
  ## We need some sample aliases that follow R naming conventions otherwise
  ## we run into errors when building upset plots
  mutate(
    alias = c(
      paste0(rep("fAD", 7), seq(1, 7)),
      paste0(rep("fAI", 8), seq(1, 8)),
      paste0(rep("wt", 9), seq(1, 9))
    )
  )
metadata %>%
  kable(
    align = "lllllll"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  )
```

```{r}
fad <- metadata$alias %>%
  str_detect("fAD")
fai <- metadata$alias %>%
  str_detect("fAI")
wt <- metadata$alias %>%
  str_detect("WT")
```

## Differential Expression

```{r}
topDE <- readRDS(here("files/toptable_raw.rds")) %>%
  set_names(c("fAD", "fAI"))
```

## GeneiASE

```{r}
files <- list.files(
  "/hpcfs/users/a1647910/210408_psen1_fADfAI_snv/15_geneiASE/output",
  full.names = TRUE
)
samples <- basename(files) %>%
  str_remove(".static.pval.tsv")
drChrs <- seq(1:25)
```

```{r}
topASE <- lapply(files, function(x){
  sample <- basename(x) %>%
    str_remove(".static.pval.tsv")
  read_tsv(x) %>%
    mutate(
      sig = fdr < 0.05,
      sample = sample
    ) %>%
    left_join(metadata[,c("sample", "alias", "genotype")]) %>%
    dplyr::arrange(fdr) %>%
    mutate(ASE = fdr < 0.05)
}) %>%
  set_names(metadata$alias[match(samples, metadata$sample)]) %>%
  .[metadata$alias]
```

```{r}
nASE <- sapply(topASE, function(x){
  dplyr::filter(x, ASE) %>%
    nrow()
})
```

# Enrichment testing

## Fisher's exact test

The first and most simple approach to testing for overall enrichment of DE genes
showing ASE is Fisher's exact test.
2x2 contingency tables were constructed for each mutant genotype (fAD and fAI).
Genotype fAD showed `r nrow(dplyr::filter(topDE$fAD, DE))` DE genes in 
comparison to WT, while genes presenting static ASE ranged from 
`r min(nASE[fad])` to `r max(nASE[fad])` across `r length(fad)` samples.
Genotype fAI showed `r nrow(dplyr::filter(topDE$fAI, DE))` DE genes in
comparison to WT, while genes presenting static ASE ranged from
`r min(nASE[fai])` to `r max(nASE[fai])` across `r length(fai)` samples.

```{r}
fisherTabs <- topASE[str_detect(names(topASE), ("fAD|fAI"))] %>%
  lapply(function(x){
    geno <- unique(x$alias)
    if (str_detect(geno, "fAD")) {
      dat <- topDE$fAD[,c("gene_id", "DE")] %>%
        left_join(x[,c("feat", "ASE")], by = c("gene_id" = "feat")) %>%
        mutate(
          DE = ifelse(DE == TRUE, "DE", "Not DE"),
          ASE = ifelse(ASE == TRUE, "ASE", "No ASE"),
          ASE = ifelse(is.na(ASE), "No ASE", ASE)
        )
      tab <- table(dat$DE, dat$ASE)
      return(tab)
    }
    if (str_detect(geno, "fAI")) {
      dat <- topDE$fAI[,c("gene_id", "DE")] %>%
        left_join(x[,c("feat", "ASE")], by = c("gene_id" = "feat")) %>%
        mutate(
          DE = ifelse(DE == TRUE, "DE", "Not DE"),
          ASE = ifelse(ASE == TRUE, "ASE", "No ASE"),
          ASE = ifelse(is.na(ASE), "No ASE", ASE)
        )
      tab <- table(dat$DE, dat$ASE)
      return(tab)
    }
  })
```

Concerns with statistical power for Fisher's exact test were assessed due to
the small number of DE genes for both genotypes.
This was performed with the `power.fisher.test()` function using 1000
simulations for each sample.

```{r}
set.seed(17)
fisherSim <- sapply(fisherTabs, function(x){
  p1 <- x[1,1] / (x[2,1])
  n1 <- (x[1,1] + x[2,1])
  p2 <- x[1,2] / (x[2,2])
  n2 <- (x[1,2] + x[2,2])
  power.fisher.test(p1, p2, n1, n2, alpha = 0.05, nsim = 1000)
})
```

```{r}
fisherSim %>%
  enframe(name = "Sample", value = "Power") %>%
  left_join(metadata, by = c("Sample" = "alias")) %>%
  ggplot(aes(Sample, Power)) +
  geom_bar(aes(fill = genotype), stat = "identity") +
  labs(fill = "Genotype")
```

Clearly Fisher's exact test lacks power to detect enrichment of DE genes
presenting ASE.
The test was stil performed as an initial insight, but other methods must be
considered to make proper conclusions.

```{r}
fisherTest <- lapply(fisherTabs, fisher.test)
fisherP <- sapply(fisherTest, function(x){x$p.value})
```

```{r}
fisherP %>%
  enframe(name = "Sample", value = "p") %>%
  left_join(metadata, by = c("Sample" = "alias")) %>%
  ggplot(aes(Sample, -log10(p))) +
  geom_bar(aes(fill = genotype), stat = "identity") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  labs(fill = "Genotype")
```

## Gene Set Enrichment Analysis

### Non-directional

