---
title: "GeneiASE methods"
author: "Lachlan Baer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    fig_width: 8
    fig_height: 6
    fig_align: "center"
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  fig.align = "center"
)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

# Setup 

```{r packages}
library(getopt)
library(here)
library(binom)
library(VGAM)
```

# Intro

The following analysis involves a recreation of methods implemented by
`GeneiASE` software ((GitHub link)[https://github.com/edsgard/geneiase])
to get a better understanding of the intermediary steps.
The motivation for this stems from noticing strange p-value estimation 
patterns from the results produced by static ASE analyis with `GeneiASE`.
In particular, some p-values were estimated at 0 as opposed to a float value.
The following code replicates the `GeneiASE` methods for statuc ASE and allows 
for a closer inspection of intermediary values.

# Start

## Static ASE functions

These functions are located in the `geneiase/lib/static.lib.R` file, and have
been copied directly.

```{r}
read.data <- function(in.file){
  if(length(grep('\\.rds$', in.file)) > 0){
    vars.filt = readRDS(in.file)
  }else{
    colClasses = c(rep('character', 2), rep('integer', 2))
    vars.filt = read.table(in.file, sep = '\t', header = TRUE, stringsAsFactors = FALSE, colClasses = colClasses)
    colnames(vars.filt) = c('gene', 'snp', 'alt.dp', 'ref.dp')
    if(ncol(vars.filt) == 5){
      colnames(vars.filt)[5] = 'p.sample.null'
      vars.filt[, 5] = as.numeric(vars.filt[, 5])
    }
    if(ncol(vars.filt) == 6){ #simulated data where class is known
      colnames(vars.filt)[5:6] = c('p.sample.null', 'class')
      vars.filt[, 5] = as.numeric(vars.filt[, 5])
    }
    
  }
  
  return(vars.filt)
}
```

```{r}
get.main.pval <- function(obs.df, method = 'wilson', n.samples, nmax.vars, betabin.rho = 0, p.sample.null = 0.5, testval.col = 'z', gene.null.nvar.list = NULL){
  
  ###
  #Obs dist
  ###
  browser()
  #Adds dp col (dp = alt.dp + ref.dp) if a dp col doesnt already exist
  obs.df = add.dp.col(obs.df)
  
  #Get SNV test-stat (z)
  obs.df = get.snv.testval(obs.df, method)
  
  #Add column: n.vars
  obs.df = add.nvar.column(obs.df)
  
  #Calc gene based measures for each gene based on the snv.testval (median, mean, cv, Stouffer)
  obs.gene.df = get.gene.testval(obs.df, testval.col = testval.col) 
  
  #Add column 'nvars.bin'
  obs.gene.df = add.nvars.bin(obs.gene.df, nmax.vars)
  
  
  ###
  #Null dist
  ###
  #Generate null by binomial sampling with same depths as the snvs within each gene, that is, one null per snv combined by stouffer into one null per gene
  #(Alt: compare the frac:s directly (estimated vs sampled), then Stouffer on those p-values)
  
  #get p.sample.null
  cols = colnames(obs.df)
  if(length(which(cols == 'p.sample.null')) > 0){
    p.sample.null = 'p.sample.null'
  }
  
  if(is.null(gene.null.nvar.list)){
    liptak.s.p.nom = get.binresamp.pval(obs.df, obs.gene.df, n.samples, method, betabin.rho, p.sample.null)
    obs.gene.df = cbind(obs.gene.df, liptak.s.p.nom)
  }
  else{ #pre-calculated null dists
    
    #split and loop on nvars.bin
    obs.gene.df.nvar.list = split(obs.gene.df, obs.gene.df[, 'nvars.bin'])
    nvars.bins = names(obs.gene.df.nvar.list)
    for(j.bin in nvars.bins){
      gene.null = gene.null.nvar.list[[j.bin]]
      obs.gene.df.nvar = obs.gene.df.nvar.list[[j.bin]]
      
      #Get p-val by comparing to null
      liptak.s.p.nom = sapply(obs.gene.df.nvar[, 'liptak.s'], function(obs.val, gene.null){p.nom = length(which(gene.null >= obs.val)) / length(gene.null)}, gene.null = gene.null)
      obs.gene.df.nvar.list[[j.bin]] = cbind(obs.gene.df.nvar, liptak.s.p.nom)
    }
    
    #rbind
    obs.gene.df = as.data.frame(matrix(nrow = 0, ncol = ncol(obs.gene.df.nvar.list[[1]])))
    for(j.bin in nvars.bins){
      obs.gene.df = rbind(obs.gene.df, obs.gene.df.nvar.list[[j.bin]])
    }
  }
  
  
  #Multi-test Adjust
  p.nom.cols = colnames(obs.gene.df)[grep('p.nom', colnames(obs.gene.df))]
  for(jcol in p.nom.cols){
    p.bh = p.adjust(obs.gene.df[, jcol], method = 'BH');
    obs.gene.df = cbind(obs.gene.df, p.bh);
    colnames(obs.gene.df)[ncol(obs.gene.df)] = paste(jcol, 'bh', sep = '.');
  }                                                                                                
  
  
  #Merge with gene class annot (neg/pos)
  #Only applicabable in the case of simulated data (not real data)
  cols = colnames(obs.df)
  if(length(grep('^class$', cols) > 0)){
    gene.annot = unique(obs.df[, c('gene', 'class')])
    obs.gene.df = merge(obs.gene.df, gene.annot, by = 'gene')
  }
  
  return(obs.gene.df)
}
```

```{r}
get.snv.testval <- function(data.df, method = 'wilson', eps = 1e-16, alt.dp.col = 'alt.dp', site.dp.col = 'dp', p.binom.null = 0.5){
  
  #make ref.dp col (doesn't always exist)
  ref.dp = data.df[, site.dp.col] - data.df[, alt.dp.col]
  
  #add pseudo-count if alt.dp or ref.dp == 0
  pseudo.data.df = cbind(data.df[, site.dp.col], data.df[, alt.dp.col], ref.dp)
  colnames(pseudo.data.df) = c('dp', 'alt.dp', 'ref.dp')
  pseudo.data.df = add.pseudo(pseudo.data.df)
  
  #get dp cols
  dp = pseudo.data.df[, 'dp']
  alt.dp = pseudo.data.df[, 'alt.dp']
  ref.dp = pseudo.data.df[, 'ref.dp']
  
  if(method == 'p.binom'){ #use p-value from binom.test to calculate Z
    
    #get p-value from binom.test
    x = cbind(alt.dp, dp)
    bt.res = t(apply(x, 1, function(x, p){res = binom.test(x['alt.dp'], x['dp'], p = p, alternative = 'two.sided'); return(c(res$p.value, res$estimate, res$conf.int[1:2]));}, p = p.binom.null))
    colnames(bt.res) = c('pval', 'p', 'p.ci.lower', 'p.ci.upper')
    p = bt.res[, 'p']
    p.ci.upper = bt.res[, 'p.ci.upper']
    p.ci.lower = bt.res[, 'p.ci.lower']
    
    #log.odds. NB: abs needed since unphased data
    abs.log.odds = abs(log(p / (1 - p)))
    
    #log.odds.se
    log.odds.se = (log(p.ci.upper) - log(p.ci.lower)) / (2 * 1.96)
    
    #Z score from inv.norm(1-pval)    
    z = pval2zscore(bt.res[, 'pval'], eps)
  }
  else{ #binom.confint
    
    #p
    p = alt.dp / dp
    
    #p.CI
    ci = binom.confint(alt.dp, dp, methods = method) #Agresti-Coull no good since some estimates < 0 and >1!
    p.ci.upper = ci[, 'upper']
    p.ci.lower = ci[, 'lower']
    
    #log.odds. NB: abs needed since unphased data
    abs.log.odds = abs(log(p / (1 - p)))
    
    #log.odds.se
    log.odds.se = (log(p.ci.upper) - log(p.ci.lower)) / (2 * 1.96)
    
    #half z (since absolute value)
    z = abs.log.odds / log.odds.se
  }
  
  #bind all
  data.df = cbind(data.df, z, abs.log.odds, log.odds.se, p, p.ci.upper, p.ci.lower)
  
  return(data.df)
}
```

```{r}
get.binresamp.pval <- function(obs.df, obs.gene.df, n.samples, method = 'wilson', betabin.rho = 0, p.sample.null = 0.5){
  
  obs.df.gene.list = split(obs.df, obs.df[, 'gene'])
  genes = names(obs.df.gene.list)
  n.genes = length(genes)
  
  p.nom = vector(length = n.genes, mode = 'numeric')
  names(p.nom) = genes
  j.gene.it = 0
  for(jgene in genes){
    j.gene.it = j.gene.it + 1
    
    if((j.gene.it %% 100) == 0){
      cat(paste('j.gene: ', j.gene.it, '\n', sep = ''))
    }
    
    #Get null gene testval (liptak.stouffer of snv testvals only, now not eg cv of snv testvals returned, due to speed and size)
    obs.df.gene = obs.df.gene.list[[jgene]]
    if(is.character(p.sample.null)){
      gene.null = get.gene.null(obs.df.gene[, 'dp'], n.samples, method, obs.df.gene[, p.sample.null], betabin.rho) #vec: len: n.samples
    }else{
      gene.null = get.gene.null(obs.df.gene[, 'dp'], n.samples, method, p.sample.null, betabin.rho) #vec: len: n.samples
    }
    
    #Get p-val by comparing to null
    obs.val = obs.gene.df[which(obs.gene.df[, 'gene'] == jgene), 'liptak.s']
    p.nom[jgene] = length(which(gene.null >= obs.val)) / length(gene.null)
  }
  return(p.nom)
}
```

```{r}
get.gene.null <- function(gene.snv.dps, n.samples, method = 'wilson', p.sample.null = 0.5, betabin.rho = 0){
  
  #get null for each snv
  n.snvs = length(gene.snv.dps)
  snv.null.dist = matrix(nrow = n.samples, ncol = n.snvs)
  if(length(p.sample.null) > 1){ #vector with varying p.sample.null
    for(j.snv in 1:n.snvs){    
      null.df = get.snv.null(gene.snv.dps[j.snv], n.samples, p.sample.null[j.snv], betabin.rho) #out: n.samples alt.dp and dp
      null.df = get.snv.testval(null.df, method)
      snv.null.dist[, j.snv] = null.df[, 'z']
    }
    
  }else{ #single p.sample.null
    for(j.snv in 1:n.snvs){    
      null.df = get.snv.null(gene.snv.dps[j.snv], n.samples, p.sample.null, betabin.rho) #out: n.samples alt.dp and dp
      null.df = get.snv.testval(null.df, method)
      snv.null.dist[, j.snv] = null.df[, 'z']
    }
  }
  #get gene testval by combining snv null testvals by liptak.stouffer
  gene.null = apply(snv.null.dist, 1, liptak.stouffer) #vec: len: n.samples
  
  return(gene.null)
}
```

```{r}
get.snv.null <- function(j.dp, n.samples, p.sample.null = 0.5, betabin.rho = 0){
  
  #sample binomial, according to p = p.sample.null  
  #alt.dp = rbinom(n.samples, j.dp, p.sample.null)
  alt.dp = rbetabinom(n.samples, j.dp, p.sample.null, betabin.rho)
  dp = rep(j.dp, n.samples)
  
  data.df = cbind(alt.dp, dp)
  
  return(data.df)
}
```

```{r}
add.dp.col <- function(data.df, alt.dp.col = 'alt.dp', ref.dp.col = 'ref.dp'){
  
  cols = colnames(data.df)
  if(length(which(cols == 'dp')) == 0){
    dp = data.df[, alt.dp.col] + data.df[, ref.dp.col]
    data.df = cbind(data.df, dp)
  }
  
  return(data.df)
}
```

```{r}
add.pseudo <- function(data.df, alt.dp.col = 'alt.dp', site.dp.col = 'dp'){
  
  zero.ind = which(data.df[, alt.dp.col] == 0)
  data.df[zero.ind, alt.dp.col] = 1
  data.df[zero.ind, site.dp.col] = data.df[zero.ind, site.dp.col] + 1
  
  zero.ind = which(data.df[, 'ref.dp'] == 0)
  data.df[zero.ind, 'ref.dp'] = 1
  data.df[zero.ind, site.dp.col] = data.df[zero.ind, site.dp.col] + 1
  
  return(data.df)
}
```

## Core functions

These functions are located in the `geneiase/lib/core.lib.R` file, and have
been copied directly.

```{r}
get.gene.testval <- function(data.df, testval.col = 'z'){
  #Calc gene based measures for each gene based on the snv.testval (median, mean, cv, Stouffer)
  
  mean.s = tapply(data.df[, testval.col], data.df[, 'gene'], mean);
  median.s = tapply(data.df[, testval.col], data.df[, 'gene'], median);
  sd.s = tapply(data.df[, testval.col], data.df[, 'gene'], sd);
  cv.s = sd.s / mean.s;
  cv.s[which(sd.s == 0)] = 0 #To avoid NAs if mean also == 0 (0/0)
  
  #Liptak-Stouffer
  gene.list = split(data.df, data.df[, 'gene'])  
  liptak.s = unlist(lapply(gene.list, function(jgene.data){liptak.stouffer(jgene.data[, testval.col])})) #equal weighing 
  
  #bind all values into matrix
  gene.vals = cbind(mean.s, median.s, sd.s, cv.s, liptak.s);
  
  #add col: n.vars
  gene2nvars = table(data.df[, 'gene'])
  gene = names(gene2nvars)
  gene2nvars = cbind(as.data.frame(as.matrix(gene2nvars)), gene, stringsAsFactors = FALSE)
  colnames(gene2nvars)[1] = 'n.vars'
  gene.vals = merge(gene2nvars, gene.vals, by.x = 'gene', by.y = 'row.names')
  rownames(gene.vals) = gene.vals[, 'gene']
  
  return(gene.vals)  
}
```

```{r}
liptak.stouffer <- function(beta, w = NA, na.rm = FALSE){
  
  if(na.rm){
    beta = beta[which(!is.na(beta))]
  }
  if(is.na(w)){
    w = rep(1, length(beta))
  }
  z = sum(w * beta) / sqrt(sum(w^2))
  return(z)
}
```

```{r}
pval2zscore <- function(p, eps = 1e-16){
  #qnorm is based on Wichura's algorithm AS 241 which provides precise results up to about 16 digits
  p[which(p > (1-eps))] = 1 - eps; 
  p[which(p < eps)] = eps;
  z = qnorm(1 - p)
  
  return(z)
}
```

```{r}
fisher.combo <- function(p){
  Xsq = -2*sum(log(p))
  combo.p.nom = 1 - pchisq(Xsq, df = 2*length(p))
  return(combo.p.nom)
}
```

```{r}
add.nvars.bin <- function(vars.data, nmax.vars = 11){
  
  nvars.bin = vars.data[, 'n.vars'];
  nvars.bin[which(nvars.bin >= nmax.vars)] = nmax.vars;
  vars.data = cbind(vars.data, nvars.bin)
  
  return(vars.data)
}
```

```{r}
add.nvar.column <- function(data.df){
  
  #n.vars
  cols = colnames(data.df)
  if(length(grep('^n.vars$', cols)) == 0){
    gene2nvars = table(data.df[, 'gene'])
    gene = names(gene2nvars)
    gene2nvars = cbind(as.data.frame(as.matrix(gene2nvars)), gene, stringsAsFactors = FALSE)
    colnames(gene2nvars)[1] = 'n.vars'
    data.df = merge(data.df, gene2nvars, by = 'gene')    
  }
  
  return(data.df)
}
```

```{r}
min.vars.filter <- function(vars.filt, min.het.vars){
  gene2nvars = table(vars.filt[, 'gene']);
  genes.pass = names(gene2nvars)[which(gene2nvars >= min.het.vars)];
  vars.filt = merge(vars.filt, as.matrix(genes.pass), by.x = 'gene', by.y = 1)
  
  return(vars.filt)
}
```

```{r}
dump.tab <- function(pval.df, pval.file, sim = FALSE){
  if(length(grep('\\.rds$', pval.file)) > 0){
    saveRDS(pval.df, file = pval.file)
  }else{
    colnames(pval.df) = sub('^gene$', 'feat', colnames(pval.df))
    colnames(pval.df) = sub('^liptak.s.p.nom$', 'p.nom', colnames(pval.df))
    colnames(pval.df) = sub('^liptak.s.p.nom.bh$', 'fdr', colnames(pval.df))
    dump.cols = c('feat', 'n.vars', 'mean.s', 'median.s', 'sd.s', 'cv.s', 'liptak.s', 'p.nom', 'fdr')
    if(sim){
      dump.cols = c(dump.cols, 'class')
    }
    write.table(pval.df[, dump.cols], quote = FALSE, row.names = FALSE, sep = '\t', file = pval.file)
  }
}
```

```{r}
get.tpr.fpr <- function(obs.mat, alpha.vec, measure, feat.col = 'gene'){
  
  #classify based on measure
  gene2sig.alpha = matrix(nrow = nrow(obs.mat), ncol = length(alpha.vec), dimnames = list(obs.mat[, feat.col], alpha.vec))
  for(jalpha in alpha.vec){
    gene2sig.alpha[, as.character(jalpha)] = get.sig(obs.mat, jalpha, measure)    
  }
  
  #Merge with gene class annot
  gene.annot = unique(obs.mat[, c(feat.col, 'class')])
  gene2sig.alpha = merge(gene.annot, gene2sig.alpha, by.x = feat.col, by.y = 'row.names')  
  
  #Get fpr and sens
  fpr.sens = t(apply(gene2sig.alpha[, as.character(alpha.vec)], 2, function(call, truth){fpr = length(which(call & (truth == 'neg'))) / length(which(truth == 'neg')); sens = length(which(call & (truth == 'pos'))) / length(which(truth == 'pos')); return(c(fpr, sens));}, truth = gene2sig.alpha[, 'class']))
  colnames(fpr.sens) = c('fpr', 'tpr')
  
  return(fpr.sens)
}
```

```{r}
get.mcc <- function(obs.mat, alpha.vec, measure, feat.col = 'gene'){
  
  #classify based on measure
  gene2sig.alpha = matrix(nrow = nrow(obs.mat), ncol = length(alpha.vec), dimnames = list(obs.mat[, feat.col], alpha.vec))
  for(jalpha in alpha.vec){
    gene2sig.alpha[, as.character(jalpha)] = get.sig(obs.mat, jalpha, measure)    
  }
  
  #Merge with gene class annot
  gene.annot = unique(obs.mat[, c(feat.col, 'class')])
  gene2sig.alpha = merge(gene.annot, gene2sig.alpha, by.x = feat.col, by.y = 'row.names')  
  
  #Get Matthews correlation coef
  mcc = t(apply(gene2sig.alpha[, as.character(alpha.vec)], 2, function(call, truth){fp = length(which(call & (truth == 'neg'))); tp = length(which(call & (truth == 'pos'))); fn = length(which(!call & (truth == 'pos'))); tn = length(which(!call & (truth == 'neg'))); denom = (as.numeric(tp) + as.numeric(fp)) * (as.numeric(tp) + as.numeric(fn)) * (as.numeric(tn) + as.numeric(fp)) * (as.numeric(tn) + as.numeric(fn)); if(denom == 0){denom = 1}; matt = ((as.numeric(tp) * as.numeric(tn)) - (as.numeric(fp) * as.numeric(fn))) / sqrt(denom); return(c(matt, fp, tp, fn, tn));}, truth = gene2sig.alpha[, 'class']))
  colnames(mcc) = c('mcc', 'fp', 'tp', 'fn', 'tn')
  
  return(mcc)
}
```

```{r}
get.sig <- function(obs.mat, alpha, pval.cols){
  n.tests = nrow(obs.mat)
  sig.res = matrix(nrow = n.tests, ncol = length(pval.cols))
  colnames(sig.res) = pval.cols
  for(jcol in pval.cols){
    sig.col = rep(FALSE, n.tests)
    sig.col[which(obs.mat[, jcol] <= alpha)] = TRUE
    sig.res[, jcol] = sig.col
  }                                                                                                  
  
  #get genes which are sig in both mean and sd
  multisig = apply(sig.res[, pval.cols, drop = FALSE], 1, function(sig.vals){all(as.logical(sig.vals))});
  
  return(multisig)
}
```

## Options

```{r}
ase.type <- "static"
in.file <- "/hpcfs/users/a1647910/210408_psen1_fADfAI_snv/15_geneiASE/input/1_KB_A10.static.tsv"
out.file <- here("files/1_KB_A10.static.pval.tsv")
betabin.p <- 0.467
betabin.rho <- 0.012
n.samples <- 1e5
min.het.vars <- 2
nmax.vars <- 100
method <- 'wilson'
```

## Method

```{r}
vars.filt = read.data(in.file)
```

```{r}
vars.filt = min.vars.filter(vars.filt, min.het.vars)
```

```{r}
n.feats = length(unique(vars.filt[, 'gene']))
cat(sprintf('Input: %i features (after filtered on -m %i)\n', n.feats, min.het.vars))
```

```{r}
#Get p-value
cat('Calculating p-values for features...\n')
if(ase.type == 'static'){
  pval.df = get.main.pval(vars.filt, method, n.samples, nmax.vars, betabin.rho, betabin.p)
}else{
  pval.df = get.main.pval(vars.filt, method, n.samples, nmax.vars, betabin.rho)
}
```

